Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines

### 2013-04-02 paper/paper.tex
Changed over to alloc/tick, discussed problems with widening with compilation.
Conflicts:

	paper/paper.tex

%\title{Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines}
\title{Correctly Optimizing Abstract Abstract Machines}

\authorinfo{J. Ian Johnson \and Nicholas Labich \and Matthew Might \and David Van Horn}
           {}
           {}
%% \authorinfo{J. Ian Johnson}
%%            {Northeastern University}
%%            {ianj@ccs.neu.edu}
%% \authorinfo{Nicholas Labich}
%%            {Northeastern University}
%%            {labichn@ccs.neu.edu}
%% \authorinfo{Matthew Might}
%%            {University of Utah}
%%            {might@cs.utah.edu}
%% \authorinfo{David Van Horn}
%%            {Northeastern University}
%%            {dvanhorn@ccs.neu.edu}
\maketitle

\begin{abstract}
The technique of \emph{abstracting abstract machines} (AAM) provides a
systematic approach for deriving computable approximations of
evaluators that are easily proved sound.
%
This article contributes a complementary step-by-step process for
subsequently going from a \naive{} analyzer derived under the AAM
approach, to an efficient and correct implementation.  The end result
of the process is a two to three order-of-magnitude improvement over
the systematically derived analyzer, making it competitive with
hand-optimized implementations that compute fundamentally less precise
results.
\end{abstract}

@@

Program analysis provides sound predictive models of program behavior, but
in order for such models to be effective, they must be efficiently
computable and correct.  Past approaches to designing program analysis
have often featured abstractions that are far removed from the
original language semantics, requiring ingenuity in their construction
and effort in their verification.
%
The \emph{abstracting abstract machines} (AAM)
approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic}
to deriving program analyses provides an alternative: a systematic way
of transforming a programming language semantics in the form of an
abstract machine into a family of abstract interpreters.  It thus reduces
the burden of constructing and verifying the soundness of an abstract
interpreter.
%

@@

Unfortunately, the AAM approach yields analyzers with poor performance
relative to hand-optimized analyzers.
%
Our work takes aim squarely at this ``efficiency gap,'' and narrows it
in an equally systematic way through a number of simple steps, many of
which are inspired by run-time implementation techniques such as
laziness and compilation to avoid interpretative overhead.  Each of
these steps is proven correct, so the end result is an implementation
that is trustworthy and efficient.

In this article, we develop a systematic approach to deriving a
practical implementation of an abstract-machine-based analyzer using
mostly semantic means rather than tricky engineering. Our goal is to
empower programming language implementers and researchers to explore
and convincingly exhibit their ideas with a low barrier to entry. The
optimizations we describe are widely applicable and apparently
effective to scale far beyond the size of programs typically considered
in the recent literature on flow analysis for functional languages.

@@

\begin{figure}[t]
\small
\begin{center}
%\includegraphics[width=3.2in]{church-relative-space}
\includegraphics[width=3.2in]{church-relative-speed}
\includegraphics[width=3.2in]{church-relative-time}
\vspace{-1.5em}
\end{center}
\caption{Factor improvements over the baseline analyzer for the
  \Church{} benchmark in terms of the rate of state
  transitions and total analysis time. (Bigger is better.) Each point
  is marked with the section that introduces the optimization.}
\label{fig:churchtime}
\end{figure}

@@

The AAM approach, in essence, does the following: it takes a
machine-based view of computation and turns it into a \emph{finitary
  approximation} by bounding the size of the store.  With a limited
address space, the store must map addresses to \emph{sets} of values.
Store updates are interpreted as joins, and store dereferences are
interpreted by non-deterministic choice of an element from a set.  The
result of analyzing a program is a finite directed graph where nodes
in the graph are (abstract) machine states and edges denote machine
transitions between states.

@@

\begin{figure}[t]
\small
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{Example state graphs for the program above.  Part (a) shows
  the result of the baseline analyzer.  It has long ``corridor''
  transitions and ``diamond'' subgraphs that fan-out from
  nondeterminism and fan-in from joins.  Part (b) shows the result of
  performing nondeterminism lazily and thus avoids many of the diamond
  subgraphs.  Part (c) shows the result of abstract compilation that
  removes interpretive overhead in the form of intermediate states,
  thus minimizing the corridor transitions.  The end result is a more
  compact abstraction of the program that can be generated faster.}
\label{fig:state-graphs}
\end{figure}

%\newpage
\section{Abstract interpretation of ISWIM}
\label{sec:aam}

@@

\begin{figure}
\small
\[
\begin{array}{l@{\qquad}rcl}
\text{Expressions} & \mexp &=& \svar[^\mlab]\mvar\\
&&|& \slit[^\mlab]\mlit\\
&&|& \slam[^\mlab]\mvar\mexp\\
&&|& \sapp[^\mlab]\mexp\mexp \\
&&|& \sif[^\mlab]\mexp\mexp\mexp \\
\text{Variables}&\mvar &=& \syntax{x}\ |\ \syntax{y}\ |\ \dots\\
\text{Literals}&\mlit &=& \mnum\ |\ \mbln\ |\ \mop\\
\text{Integers}&\mnum &=& \syntax{0}\ |\ \syntax{1}\ |\ \syntax{-1}\ |\ \dots\\
\text{Booleans}&\mbln &=& \strue\ |\ \sfalse\\
\text{Operations}&\mop &=& \syntax{zero?}\ |\ \syntax{add1}\ |\ \syntax{sub1}\ |\ \dots
\end{array}
\]
\caption{Syntax of ISWIM}
\label{fig:syntax}
\end{figure}

\begin{figure}
\small
\[
\begin{array}{l@{\qquad}rcl}
\text{Values} & \mval,\mvalx{u} &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\ |\ \mkont\\
\text{States} & \mstate &=& \ev{\mexp,\menv,\msto,\mkont}\\
                       &&|& \co{\mkont,\mval,\msto}\\
                       &&|& \ap{\mval,\mval,\msto,\mkont}\\
\text{Continuations} & \mkont &=& \kmt\\
&&|& \kfn{\mval,\mkaddr}\\
&&|& \kar{\mexp,\menv,\mkaddr}\\
&&|& \kif{\mexp,\mexp,\menv,\mkaddr}\\
\text{Addresses} &\maddr,\mkaddr&\in&\Addr \\
\text{Environments} &\menv&\in& \Var \rightharpoonup \Addr\\
\text{Stores} &\msto&\in& \Addr \rightharpoonup \mathcal{P}(\Value)
\end{array}
\]
\caption{Abstract machine components}
\label{fig:domains}
\end{figure}

The semantics are defined in terms of a machine model.  The machine
components are defined in figure~\ref{fig:domains};
figure~\ref{fig:aam} defines the transition relation.  The evaluation
of a program is defined as its set of traces that arise from iterating the
machine transition relation.  The
machine is a very slight variation on a standard abstract machine for
ISWIM in ``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.  It
can be systematically derived from a definitional interpreter through
a continuation-passing style transformation and defunctionalization,
or from a structural operational semantics using the refocusing
construction of Danvy and
Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.

Compared with the standard machine semantics, this definition is
different in the following ways, which make it abstractable as a
program analyzer:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values,\footnote{
More generally, we can have stores map to any domain that forms
 a Galois connection with values, enabling $\interpdelta$ to
produce elaborate abstractions of base values (e.g., interval or octagon
abstractions). We use sets of values for a simpler exposition.
} not
  single values,

\item continuations are heap-allocated, not stack-allocated,
\item there are ``timestamps'' ($\mcntr \in \Counter$) and syntax
  labels ($\mlab$) threaded through the computation, and
\item the machine is implicitly parameterized by the functions
  $\alloc$, $\allockont$, $\tick$, $\interpdelta$, and spaces
  $\Addr$, $\Counter$ (and initial $\mcntr_0 \in \Counter$).
\end{itemize}

\begin{figure}
\small
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \{ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{array}{@{}r@{\ }c@{\ }l@{}}
\mstate &\machstep&\mstate' \text{ defined to be the following} \\
&&\text{let } \mcntr' =\tick(\mstate) \\
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mval,\msto}
\text{ if }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mlit,\msto}
\\
\ev[^\mcntr]{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
\co[^{\mcntr'}]{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\mkaddr}}
\\
&&
\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\mkaddr}}
\\
&&
\text{ where }\mkaddr = \allockont^\mcntr_\mlab(\msto,\mkont) \\
&&\phantom{\text{ where }}\msto' = \msto\sqcup[\mkaddr \mapsto \setof{\mkont}]
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep&
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep&
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\mkaddr}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\mkaddr},\mval,\msto} & \machstep&
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ if }\mkont \in \msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\strue,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\mkaddr)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\mkaddr},\sfalse,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi1,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\mkaddr)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\\
&&\text{ where } \maddr  =\alloc(\mstate) \\
&&\phantom{\text{ where }} \menv' = \menv[\mvar\mapsto\maddr] \\
&&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr\mapsto\{\mval\}] \\
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
\co{\mkont,\mval',\msto}
\text{ if } \mval'\in\interpdelta(\mop,\mval)
\end{array}
\end{gather*}
\caption{Abstract abstract machine for ISWIM}
\label{fig:aam}
\end{figure}

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \text{ the } \msto \text{ within } \mstate \\
\allockont^\mcntr_\mlab(\msto,\mkont) &=\mkaddr \mbox{ where } \mkaddr \notin \msto
\end{align*}
These functions appear to ignore $\mlab$ and $\mcntr$, but they can be
used to determinize the choice of fresh addresses. The $\sqcup$ on
stores in the figure is a point-wise lifting of $\cup$: $\msto \sqcup \msto' =
\lambda \maddr. \msto(\maddr) \cup \msto'(\maddr)$. The resulting
relation is non-deterministic in its choice of addresses, however it
must always choose a fresh address when allocating a continuation or
variable binding.  If we consider machine states equivalent up to
consistent renaming and fix an allocation scheme, this relation
defines a deterministic machine (the relation is really a function).

@@

\paragraph{Abstract interpretation} To characterize abstract
interpretation, set the implicit parameters just as above, but drop
the $\maddr \not\in \msto$ condition. The $\interpdelta$ relation takes some care
to not make the analysis run forever; a simple instantiation is a flat
abstraction where arithmetic operations return an abstract top element
$\sNum$, and $\szerohuh$ returns both $\strue$ and $\sfalse$ on
$\sNum$.  This family of interpreters is also non-deterministic in
choices of addresses, but it is free to choose addresses that are
already in use.  Consequently, the machines may be non-deterministic
when multiple values reside in a store location.

It is important to recognize from this definition that \emph{any}
allocation strategy is a sound abstract
interpretation~\cite{dvanhorn:Might2009Posteriori}.  In particular,
concrete interpretation is a kind of abstract interpretation.  So is
an interpretation that allocates a single cell into which all bindings
and continuations are stored.  The former is an abstract
interpretation that is non-computable and gives only the ground truth
of a programs behavior; the latter is an abstract interpretation
that is easy to compute but gives little information.  Useful program
analyses lay somewhere in between and can be characterized by their
choice of address representation and allocation strategy.  Uniform
\(k\)-CFA~\cite{dvanhorn:nielson-nielson-popl97}, presented next, is one such analysis.

@@

{\small
\begin{align*}
\Counter &= \Label^* \\
\mtcntr &= \epsilon \\
\alloc(\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &= \mlab\mcntr \\
\alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
\allockont^\mcntr_\mlab(\msto,\mkont) &= \mlab\mcntr \\
\tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
\tick(\co{\kar[^\mcntr]{\mexp,\menv,\mkaddr},\mval,\msto}) &= \mcntr \\
\tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
  \kpush[_0]{\mcntr} &= \kpush[_k]{\mtcntr} = \mtcntr \\
  \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
\end{align*}}
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

\section{From machine semantics to baseline analyzer}
\label{sec:baseline}

The uniform $k$-CFA allocation strategy would make $\traces$ in figure
\ref{fig:aam} a computable abstraction of possible executions, but one
that is too inefficient to run, even on small examples.  Through this
section, we explain a succession of approximations to reach a more
appropriate baseline analysis.
%
We ground this path by first formulating the analysis in terms of a
classic fixed-point computation.

@@

Conceptually, the AAM approach calls for computing an analysis as a
graph exploration: (1) start with an initial state, and (2) compute
the transitive closure of the transition relation from that state. All
visited states are potentially reachable in the concrete, and all
paths through the graph are possible traces of execution.

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\mstate_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\mstate_0} : \wp(\State) \times \wp(\State\times\State) \to \wp(\State) \times \wp(\State\times\State)\text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
\begin{align*}
  F_{\mstate_0}(V,E) &= (\{ \mstate_0 \} \cup V', E') \\
    E' &= \setof{ (\mstate,\mstate') \mid \mstate \in V \text{ and } \mstate \machstep \mstate'} \\
    V' &= \setof{ \mstate' \mid (\mstate,\mstate') \in E'}
\end{align*}
Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
% \begin{equation*}
 $\eval(\mexp) = \mathrm{lfp}(F_{\mstate_0})\text{,}$
%\end{equation*}
where $\mstate_0 = \ev[^\mtcntr]{\mexp, \varnothing, \varnothing, \kmt}$.

The possible traces of execution tell us the most about a program, so
we take $\traces(\mexp)$ to be the (regular) set of paths through the
computed graph. We elide the construction of the set of edges in this paper.

@@

In the next subsection, we fix this with store widening to reach polynomial
(albeit of high degree) complexity.
%
This widening effectively lifts the store out of individual states to create
a single, global shared store for all.

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as the
application of a widening operator during the fixed-point iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
program.
%
Thus, complexity drops from intractable exponentiality to a merely
daunting polynomial.

>>>>

\title{Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines}

\authorinfo{J. Ian Johnson}
           {Northeastern University}
           {ianj@ccs.neu.edu}
\authorinfo{Matthew Might}
           {University of Utah}
           {might@cs.utah.edu}
\authorinfo{David Van Horn}
           {Northeastern University}
           {dvanhorn@ccs.neu.edu}

\maketitle

\begin{abstract}
Abstracting abstract machines has been proposed as a
lightweight approach to designing sound and computable program
analyses.  The approach derives abstract interpreters from
existing machine semantics and has been applied to a variety of
languages with features widely considered difficult to analyze.
Although sound analyzers are straightforward to build under this approach,
they are also prohibitively inefficient.

This article contributes a step-by-step process for going from a
\naive{} analyzer derived under the abstracting abstract machine
approach to an efficient program analyzer.  The end result of the
process is a two to three order-of-magnitude improvement over the
systematically derived analyzer, making it competitive with
hand-optimized implementations that compute fundamentally less precise
results.
\end{abstract}

@@

The \emph{abstracting abstract machines} (AAM)
approach~\cite{dvanhorn:VanHorn2011Abstracting,dvanhorn:VanHorn2012Systematic}
to deriving program analyses provides a systematic way of transforming a
programming language semantics in the form of an abstract machine into a family
of abstract interpreters.
%
The approach parameterizes these families with policies for regulating analytic
precision.
%
While flexible and robust, the AAM approach unfortunately yields analyzers with
poor performance relative to hand-optimized analyzers.
%
Our work takes aim squarely at this ``efficiency gap,'' and narrows it in an
equally systematic way.

@@

However, there is a considerable weakness with the approach: an
analyzer designed and implemented by following the AAM recipe is
prohibitively inefficient without both further approximation and
further implementation effort.

In this article, we develop a systematic approach to deriving a
feasible implementation of an abstract-machine-based analyzer.

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=3.2in]{church-relative-space}
\includegraphics[width=3.2in]{church-relative-speed}
\includegraphics[width=3.2in]{church-relative-time}
\vspace{-1.5em}
\end{center}
\caption{Factor improvements over the baseline analyzer for the
  \Church{} benchmark in terms of peak memory usage, the rate of state
  transitions, and total analysis time. (Bigger is better.) Each point
  is marked with the section that introduces the optimization.}
\label{fig:churchtime}
\end{figure}

@@

\begin{figure}[t]
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{Example state graphs for the program above.  Part (a) shows
  the result of the baseline analyzer.  It has long ``corridor''
  transitions and ``diamond'' subgraphs that fan-out from
  nondeterminism and fan-in from joins.  Part (b) shows the result of
  performing nondeterminism lazily and thus avoids many of the diamond
  subgraphs.  Part (c) shows the result of abstract compilation that
  removes interpretive overhead in the form of intermediate states,
  thus minimizing the corridor transitions.  The end result is a more
  compact abstraction of the program that can be generated faster.}
\label{fig:state-graphs}
\end{figure}

The AAM approach, in essence, does the following: it takes a
machine-based view of computation and turns it into a \emph{finitary
  approximation} by bounding the size of the store.  With a limited
address space, the store must map addresses to \emph{sets} of values.
Store updates are interpreted as joins, and store dereferences are
interpreted by non-deterministic choice of an element from a set.  The
result of analyzing a program is a finite directed graph where nodes
in the graph are machine states and edges denote machine
transitions between states.

@@

%\newpage
\section{An analyzer for ISWIM}
\label{sec:aam}

@@

\begin{figure}
\[
\begin{array}{l@{\qquad}rcl}
\text{Expressions} & \mexp &=& \svar[^\mlab]\mvar\\
&&|& \slit[^\mlab]\mlit\\
&&|& \slam[^\mlab]\mvar\mexp\\
&&|& \sapp[^\mlab]\mexp\mexp \\
&&|& \sif[^\mlab]\mexp\mexp\mexp \\
\text{Variables}&\mvar &=& \syntax{x}\ |\ \syntax{y}\ |\ \dots\\
\text{Literals}&\mlit &=& \mnum\ |\ \mbln\ |\ \mop\\
\text{Integers}&\mnum &=& \syntax{0}\ |\ \syntax{1}\ |\ \syntax{-1}\ |\ \dots\\
\text{Booleans}&\mbln &=& \strue\ |\ \sfalse\\
\text{Operations}&\mop &=& \syntax{zero?}\ |\ \syntax{add1}\ |\ \syntax{sub1}\ |\ \dots
\end{array}
\]
\caption{Syntax of ISWIM}
\label{fig:syntax}
\end{figure}

\begin{figure}
\[
\begin{array}{l@{\qquad}rcl}
\text{Values} & \mval,\mvalx{u} &=& \clos{\mvar,\mexp,\menv}\ |\ \mlit\ |\ \mkont\\
\text{States} & \mstate &=& \ev{\mexp,\menv,\msto,\mkont}\\
                       &&|& \co{\mkont,\mval,\msto}\\
                       &&|& \ap{\mval,\mval,\msto,\mkont}\\
\text{Continuations} & \mkont &=& \kmt\\
&&|& \kfn{\mval,\maddr}\\
&&|& \kar{\mexp,\menv,\maddr}\\
&&|& \kif{\mexp,\mexp,\menv,\maddr}\\
\text{Addresses} &\maddr&\in&\Addr \\
\text{Environments} &\menv&\in& \Var \rightharpoonup \Addr\\
\text{Stores} &\msto&\in& \Addr \rightharpoonup \mathcal{P}(\Value)
\end{array}
\]
\caption{Abstract machine components}
\label{fig:domains}
\end{figure}

The semantics are defined in terms of a machine model.  The machine
components are defined in figure~\ref{fig:domains} and
figure~\ref{fig:aam} defines the transition relation.  The evaluation
of a program is defined as its traces given by the machine transition relation.  The
machine is a very slight variation on a standard abstract machine for
ISWIM in ``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.  It
can be systematically derived from a definitional interpreter through
a continuation-passing style transformation and defunctionalization,
or from a structural operational semantics using the refocusing
construction of Danvy and
Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.

Compared with the standard machine semantics, this definition is
different in the following ways, which make it abstractable as a
program analyzer:
\begin{itemize}
\item the store maps addresses to \emph{sets} of values, not
  single values,
\item continuations are heap-allocated, not stack-allocated,
\item there are ``timestamps'' (written $\mcntr$) and syntax
  labels ($\mlab$) threaded through the computation, and
\item the machine is implicitly parameterized by the functions
  $\mathit{alloc}$, $\mathit{tick}$,
  $\mathit{push}$, $\interpdelta$, and spaces
  $\Addr$, $\Counter$ (and initial $\mcntr_0 \in \Counter$).
\end{itemize}

More generally, we can make stores map to a domain that forms a galois
connection with sets of values, so that $\interpdelta$ can give more
elaborate abstractions of base values (e.g., with the interval
abstraction, $\{0,[0,5]\} \equiv \{[0,5]\}$). We stick with sets for
simpler exposition.

\begin{figure}
\begin{gather*}
\begin{align*}
\traces(\mexp) &= \{ \ev[^{\mtcntr}]{\mexp,\varnothing,\varnothing,\kmt} \multimachstep \mstate \} \text{ where }
\end{align*}
\\[2mm]
\begin{array}{@{}r@{\ }c@{\ }l@{}}
&&\mcntr' =\tick(\mcntr) \\
&&\maddr  =\alloc(\mstate) \\
%% EVAL
\ev{\svar\mvar,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mval,\msto}
\text{ if }\mval \in \msto(\menv(\mvar))
\\
\ev{\slit\mlit,\menv,\msto,\mkont} &\machstep&
\co{\mkont,\mlit,\msto}
\\
\ev[^\mcntr]{\slam\mvar\mexp,\menv,\msto,\mkont} &\machstep&
\co[^{\mcntr'}]{\mkont,\clos{\mvar,\mexp,\menv},\msto}
\\
\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi{0},\menv,\msto',\kar[_\mlab^\mcntr]{\mexpi{1},\menv,\maddr}}
\\
&&
\text{ where }\maddr,\msto' = \mathit{push}^\mcntr_\mlab(\msto,\mkont)
\\
\ev[^\mcntr]{\sif[^\mlab]{\mexpi0}{\mexpi1}{\mexpi2},\menv,\msto,\mkont} &\machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto',\kif[^\mcntr]{\mexpi1,\mexpi2,\menv,\maddr}}
\\
&&
\text{ where }\maddr,\msto' = \mathit{push}_\mlab^\mcntr(\msto,\mkont)
\\[2mm]
%% CONTINUE
\co{\kmt,\mval,\msto} &\machstep&
\ans{\msto,\mval}
\\
\co{\kar[^\mcntr_\mlab]{\mexp,\menv,\maddr_\mkont},\mval,\msto} & \machstep&
\ev[^\mcntr]{\mexp,\menv,\msto,\kfn[^\mcntr_\mlab]{\mval,\maddr_\mkont}}
\\
\co{\kfn[^\mcntr_\mlab]{{\mvalx{u}},\maddr_\mkont},\mval,\msto} & \machstep&
\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont,\msto}
\text{ if }\mkont \in \msto(\maddr_\mkont)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\strue,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi0,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\maddr_\mkont)
\\
\co{\kif[^\mcntr]{\mexpi0,\mexpi1,\menv,\maddr_\mkont},\sfalse,\msto} & \machstep&
\ev[^{\mcntr'}]{\mexpi1,\menv,\msto,\mkont}
\text{ if }\mkont\in\msto(\maddr_\mkont)
\\[2mm]
%% APPLY
\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont} & \machstep&
\ev[^{\mcntr'}\!]{\mexp,\menv',\msto',\mkont}
\\
&&\text{ where } \menv' = \menv[\mvar\mapsto\maddr] \\
&&\phantom{\text{ where }} \msto' = \msto\sqcup[\maddr\mapsto\{\mval\}] \\
\\
\ap[^\mcntr_\mlab]{\mop,\mval,\msto,\mkont} & \machstep&
\co{\mkont,\mval',\msto}
\text{ if } \mval'\in\interpdelta(\mop,\mval)
\end{array}
\end{gather*}
\caption{Abstract abstract machine for ISWIM}
\label{fig:aam}
\end{figure}

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\push_\mlab^\mcntr(\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
\mbox{ where }\maddr \notin\msto
\\
\alloc(\mstate) &= \maddr \mbox{ where } \maddr \notin \mstate.\msto
\end{align*}
where $\sqcup$ on stores is a point-wise lifting of $\sqcup$: $\msto
\sqcup \msto' = \lambda \maddr. \msto(\maddr) \sqcup
\msto'(\maddr)$. These functions appear to ignore $\mlab$ and
$\mcntr$, but they can be used to determinize the choice of fresh
addresses.  The resulting relation is non-deterministic in its choice
of addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming and fix an allocation
scheme, this relation defines a deterministic machine (the relation is
really a function).

@@

\paragraph{Abstract interpretation} To characterize abstract
interpretation, set the implicit parameters just as above, but drop
the $\maddr \not\in \msto$ condition. $\interpdelta$ takes some care
to not make the analysis run forever; a simple instantiation is a flat
abstraction that returns an abstract top element $\mathit{Integer}$ as
the result of arithmetic operations.  This family of interpreters is
also non-deterministic in choices of addresses, but it is free to
choose addresses that are already in use.  Consequently, the machines
may be non-deterministic when multiple values reside in a store
location.

It is important to recognize from this definition that \emph{any}
allocation strategy is an abstract
interpretation~\cite{dvanhorn:Might2009Posteriori}.  In particular,
concrete interpretation is a kind of abstract interpretation.  So is
an interpretation that allocates a single cell into which all bindings
and continuations are stored.  On the one hand is an abstract
interpretation that is non-computable and gives only the ground truth
of a programs behavior; on the other is an abstract interpretation
that is easy to compute but gives little information.  Useful program
analyses lay somewhere in between and can be characterized by their
choice of address representation and allocation strategy.  Uniform
\(k\)-CFA~\cite{dvanhorn:nielson-nielson-popl97} is one such analysis.

@@

\begin{align*}
\Counter &= \Label^* \\
\mtcntr &= \epsilon \\
\alloc(\ev[^\mcntr]{\sapp[^\mlab]{\mexpi0}{\mexpi1},\menv,\msto,\mkont} &= \mlab\mcntr \\
\alloc(\ap[^\mcntr_\mlab]{\clos{\mvar,\mexp,\menv},\mval,\msto,\mkont}) &= \mvar\kpush[_k]{\mlab\mcntr} \\
\push_\mlab^\mcntr(\msto,\mkont) &=
  \mlab\mcntr,\msto\sqcup[\mlab\mcntr\mapsto\{\mkont\}] \\
\tick(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont}) &= \mcntr \\
\tick(\co{\kar[^\mcntr]{\mexp,\menv,\maddr},\msto,\mval}) &= \mcntr \\
\tick(\ap[^\mcntr_\mlab]{\mvalx{u},\mval,\mkont}) &= \kpush[_k]{\mlab\mcntr} \\
  \kpush[_0]{\mcntr} &= \kpush[_k]{\mtcntr} = \mtcntr \\
  \kpush[_{k+1}]{\mlab\mcntr} &= \mlab\kpush[_k]{\mcntr} \\
\end{align*}
The \(\lfloor\cdot\rfloor_k\) notation denotes the truncation of a list
of symbols to the leftmost \(k\) symbols.

@@

\section{Reduction semantics to baseline analyzer}
\label{sec:baseline}

The uniform $k$-CFA allocation strategy would make $\traces$ in figure
\ref{fig:aam} a computable abstraction of possible executions, but not an
efficient one. This is not the strategy that AAM, nor we, recommend. Through
this section, we will explain a succession of approximations
to reach the baseline analysis.  We'll compare performance at each stage to
identify the criticality of each choice. 
%
We ground this journey by first formulating the analysis in terms of a classic
fixed-point computation.

@@

Conceptually, the AAM approach calls for computing an analysis as a graph
exploration: (1) start with an initial state, and (2) compute the transitive
closure of the transition relation from that state.

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\varsigma_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\varsigma_0} : \mathcal{P}(\State) \to \mathcal{P}(\State)
 \text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
\begin{equation*}
  F_{\varsigma_0}(S) = \{ \varsigma_0 \} \cup \{ \varsigma' \mathrel{|} \varsigma \in S \text{ and } \varsigma \machstep \varsigma' \}\text. 
\end{equation*}
Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
\begin{equation*}
 \eval(e) = \mathrm{lfp}(F_{\varsigma_0})\text{,}
\end{equation*}
where $\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}$.

{\bf FIXME} use $\traces$

@@

In the next subsection, we'll fix this with a widening and reach polynomial
(albeit of a high degree) complexity.
%
This widening effectively lifts the store out of individual states to create
a single, global shared store for all.

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as the
application of a widening operator during the fixed-point iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
program.
%
Thus, in one step, complexity drops from intractable exponentially to a merely
daunting polynomial.
